---
title: "ggdemo"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ggplot2 and dplyr

ggplot2, dplyr, and reshape2 are packages that don't come with R automatically. Instead, 
they'll need to be downloaded. Use the below code if you haven't ever run it
before:

>install.packages("ggplot2")

>install.packages("dplyr")

>install.packages("reshape2")

These packages are part of the so-called ["tidyverse"](https://www.tidyverse.org/) 
which collects packages that share the same underlying design philosophy. This 
makes it really easy to send data between different packages within the
tidyverse, but also makes it harder to send things in and out.

## Loading the libraries and LOBSTAHS data

I'll be demoing a lot of the ggplot functions using our LOBSTAHS data, freshly
produced by Savio.

```{r loading}
library(dplyr)
library(ggplot2)
library(reshape2)


#Read in the csv file from the working directory
LOBdata <- read.csv("LOBsetPos.csv")

#Check that it all looks pretty
head(LOBdata)

#Save the names of the dataset as a variable
nms <- names(LOBdata)
```

Looks good! We've got all the columns we need. I've created a data dictionary
for it below and bolded the columns that I use the most.

- X: Line number in xcms
- **peakgroup_mz**: the m/z value of the peak group. This is essentially the mass
of the lipid
- peakgroup_mzmin: the m/z value at which the peak starts
- peakgroup_mzmax: the m/z value at which the peak ends
- **peakgroup_rt**: the retention time of the peak group. This is how polar/non-
polar the lipid is.
- peakgroup_rtmin: the retention time at which the peak starts
- peakgroup_rtmax: the retention time at which the peak ends
- npeaks: the number of nicely defined peaks. Usually equal to the number of
samples in the data set (18)
- X0_uM_H2O2: the number of samples in the 0 uM hydrogen peroxide treatment
- X150_uM_H2O2: the number of samples in the 150 uM hydrogen peroxide treatment
- X30_uM_H2O2: the number of samples in the 30 uM hydrogen peroxide treatment
- **Orbi_whatever**: the intensity recorded for a given lipid for a given 
sample. This tells us how much of a given lipid is present.
- xcms_peakgroup: used for bug testing
- isotopes: describes the isotopes found in the data set and possible overlaps
- CAMERA_pseudospectrum: used for bug testing
- LOBdbase_frag_id: used by LOBSTAHS when working with MS<sup>n</sup> data
- LOBdbase_exact_parent_neutral_mass: the exact mass of the parent lipid, 
determined by subtracting the mass of the adduct
- **LOBdbase_mz**: the mass of the lipid plus the mass of the adduct. 
    - This one is bolded because we use it in Maven to look for as specific lipid.
- **lipid_class**: the lipid class (see Fahy 2005) that the lipid belongs to.
    - IP_DAG, TAG, pigment, etc.
- **species**: the lipid species the lipid belongs to
    - PC, PE, PG, MGDG, SQDG, etc.
- major_adduct: determined by the lipid class, provides the major adduct
that the lipid will form
    - NOTE: LOBSTAHS *only* uses this major adduct, and minor adducts are discarded
- FA_total_no_C: the total number of carbons in the fatty acid chains of the
lipid
- FA_total_no_DB: the total number of double bonds in the lipid
- degree_oxidation: the degree of oxidation for a given lipid
- elem_formula: the elemental formula of the given lipid compound. Usually not
unique due to regioisomers
- **compound_name**: the name of the compound. Includes oxidation information
- **the C series**: measures of confidence in the sample annotation. See Collins
et al. 2016
    - C1: Denotes a good, high confidence annotation
    - C1x: Denotes a good, high confidence annotation
    - C2a: Denotes a good, high confidence annotation
    - C2b: Denotes a good, high confidence annotation
    - C3r: Denotes regioisomers
    - C3f: Compound's m/z was identical to a competing assignment
    - C3c: Compounds isobaric to this were also present
    - C4: Lipid species was *not* formed by a major adduct and was thus discarded
    - C5: Lipid species was *not* formed by a major adduct and was thus discarded
    - C6a: ???
    - C6b: ???
- casecodes: lists the C series that the compound was flagged with
- LOBdbase_ppm_match: describes how closely the observed peak matches the
expected. Calculated by the formula ppm= (m/z_{obs}-m/z_{database})/m/z_{database}*10^6
- match_ID: Line number in LOBSTAHS

Phew. Let's simplify our LOBSTAHS dataset by just grabbing the columns and rows 
we wantand organizing them more nicely.

## Cleaning up the data

Now, the data here is *everything*. But for exploratory and introductory
analyses, we don't want to have to reference thick books and consolidate
retention times - we'd rather just use the high-confidence annotations.
To do that, I'll be filtering out the rows that belong to either C1, C1x, C2a or
C2b. The command here is filter(), from dplyr, which lets me grab rows based
on the value of a given row. I'll do this a couple different ways.

```{r cleaningrows}
#The long, terrible way: throw out each non-high-confidence annotation one at a time
clean_LOBdata <- filter(LOBdata, LOBdata$C3c!=1)
clean_LOBdata <- filter(clean_LOBdata, C3f!=1)
clean_LOBdata <- filter(clean_LOBdata, C3r!=1)
clean_LOBdata <- filter(clean_LOBdata, C4!=1)
clean_LOBdata <- filter(clean_LOBdata, C5!=1)
clean_LOBdata <- filter(clean_LOBdata, C6a!=1)
clean_LOBdata <- filter(clean_LOBdata, C6b!=1)


#The logical way
clean_LOBdata <- filter(LOBdata, C3c!=1 & C3f!=1 & C3r!=1 & C4!=1 & C5!=1 &
                          C6a!=1 & C6b!=1)
    #The ampersand above is used as an "and" logic operator

#The *almost* dplyr way
clean_LOBdata <- filter(mutate(LOBdata, ConfVals=C3c+C3f+C3r+C4+C5+C6a+C6b),
                        ConfVals==0)

#The dplyr way
clean_LOBdata <- LOBdata %>%
  mutate(ConfVals=C3c+C3f+C3r+C4+C5+C6a+C6b) %>%
  filter(ConfVals==0)
    #The %>% is called a "pipe" and passes output from one function to the 
    #other. This makes the code that would normally be nested much more readable
```

The above code reduces our data set from two thousand compounds down to 223, but
we're *very* sure all of those are what we think they are. Now that the rows are
nice, we can clean up the columns.

Here, I'll be using the "select" command from the dplyr package. It lets me grab
columns by either number or name.

```{r cleaningcols}
clean_LOBdata <- select(clean_LOBdata, 
                        c("match_ID", "lipid_class", "compound_name", "species",
                          "peakgroup_mz", "peakgroup_rt", 12:29))
    #Notice that I'm able to use both column names and the column index to grab
    #these values, rather than needing to list out the Orbi names
```

Let's take a look at our shiny clean data:

```{r lookabout}
head(clean_LOBdata)
```

Lovely! There's one more step we should take, and that's the conversion of this
"wide" data set to "long" format. ggplot only likes data in "long" format, so
we might as well take care of this now. Right now, each of our samples has its
own column. What we'd like is for that all to fit into two simple columns:
one called "sample" and one called "value". Check out [this link](https://en.wikipedia.org/wiki/Wide_and_narrow_data) if that doesn't make 
sense. There's also a few discussions online about it with better examples.

```{r reshape}
long_LOBdata <- clean_LOBdata %>%
  melt(id=c("match_ID", "lipid_class", "compound_name", "species", 
            "peakgroup_mz", "peakgroup_rt"), 
       variable.name="sample", value.name = "intensity")

head(long_LOBdata)
```

Finally, let's break up the new "Sample" column into something more useful. 
Right now, R doesn't know that we have samples from 0, 8, and 24 hours as well
as 3 different treatments. We'll use the mutate() function here to create new
columns at the end of the data frame to hold this new info.

```{r expand}
long_LOBdata <- long_LOBdata %>%
  mutate(treatment=paste(regmatches(sample, regexpr("[[:digit:]]+", sample)), "uM"),
         time_point=sub("_", "", paste(regmatches(sample, regexpr("_[[:digit:]]+", sample)), "hours")),
         sample=paste("Orbi_", regmatches(sample, regexpr("04[[:digit:]]+", sample))))
    #Use regular expressions to extract the relevant info from the sample name

head(long_LOBdata)
```

Excellent. Now that's a data set we can work with!

## ggplotting

ggplots are built in a layered format, just like base plots, but they're all 
written on the same line of code and connected via "+" signs.

All ggplots consist of 3 things - a call to ggplot(), a "geom" that determines
what shape the plot will be, and "aes()"thetics that actually fill in the data.
For example, a "skeleton" ggplot call might be:

>ggplot() + geom_something(data=, aes(x= ))

However, this format is flexible. Data could be called by the ggplot, or the
aes call could go there. In the future, this will make a difference, but for
right now it's flexible.

Some common ggplots are shown below:

```{r ggbasics, fig.width=4, fig.height=3}
#Create some fake data in long format
phytoplankton <- rep(c("Diatom", "Cocco", "Dino"), each=24)
lipid_class <- rep(rep(c("IP_DAG", "TAG", "FA"), each=8), 3)
intensity <- rep(c(20, 10, 30), each=24)+rnorm(72)
df <- data.frame(phytoplankton, lipid_class, intensity)

    #Always a good place to start
ggplot() + geom_point(data = df, aes(x=phytoplankton, y=intensity))

    #Jittering spreads out the data to avoid overlap
ggplot(data = df) + geom_jitter(aes(x=phytoplankton, y=intensity))

    #Boxplots are a nice way to present this data
ggplot(data = df, aes(x=phytoplankton, y=intensity)) + geom_boxplot()
```

However, ggplot also has a lot of customization options, just like base plot. 
Let's add some color to these - note that ggplot automatically slices the 
data for us, rather than having to make several separate calls to boxplot in
different colors

```{r prettygg}
ggplot(data = df) + geom_boxplot(aes(x=phytoplankton, 
                                     y=intensity, 
                                     color=lipid_class))
```

Sometimes, it's a good idea to "facet" your data, or present it in entirely 
separate regions within the plot area. To do this, we add (+) the facet_wrap()
layer. There's a tilde (~) before the thing you're wrapping around, don't forget!

```{r prettygg2}
ggplot(data = df) + 
  geom_boxplot(aes(x=phytoplankton, y=intensity, color=lipid_class)) +
  facet_wrap(~phytoplankton, scales = "free_x")

ggplot(data = df) + 
  geom_boxplot(aes(x=phytoplankton, y=intensity, color=lipid_class)) +
  facet_wrap(~lipid_class)
```

Note that slicing the data in this way lets us compare two different things. In 
the first one, we're emphasizing the similarity of the lipid classes within the
organisms. In the second, we're emphasizing how each phytoplankter has a 
significantly different lipid signature from the others.

As with base plot, you can control the axes, text, and titles, but here it's 
all done with the addition of a new layer by "+" rather than a second line.

```{r gaudygg}

ggplot() + geom_jitter(data=df, aes(x=phytoplankton, y=intensity, color=lipid_class)) +
  ylab("") + #remove y axis
  xlab("This is the x axis") + #add x axis label
  ggtitle("This is the title") + #add a title
  scale_x_discrete(breaks=c("Diatom", "Cocco", "Dino"), #move x axis around
                   labels=c("Bacillariophyceae", "Coccolithophore", "Dinoflagellate"))+
  scale_y_continuous(breaks = seq(10,30,1)) + #make finer slices in the background
  theme(axis.title.x = element_text(face="bold", colour="#990000", size=20)) #change axis text color
```

You may have also noticed that ggplot creates a legend on the side by default -
which is great, unless it becomes another thing to change.

```{r legendgg}
ggplot(data = df, aes(x=phytoplankton, y=intensity, color=lipid_class)) + 
  geom_point() +
  scale_color_discrete(breaks=c("IP_DAG", "TAG", "FA"),#change the order
                       name="This is the legend") + #change the title
  theme(legend.title = element_text(colour="blue", size=16, face="bold"),
        legend.text = element_text(color = "skyblue", size = 8), #change text colors
  legend.background = element_rect(fill="gray90", size=.5, linetype="dotted"), #add a background
  legend.position = "top") #move it to the top
```

## Some LOBSTAHS plots

Let's recreate the plots we covered in the demo session. We've already got our 
dataset cleaned up and ready to go, so we can jump right in to graphing it.

```{r LOBbase}
ggplot(data = long_LOBdata) + geom_point(aes(x=peakgroup_rt, y=peakgroup_mz,
                                          color=species))

```

Looks pretty busy. Let's clear out all the lipid species that Collins et al. 
also didn't use - leaving behind PC, PE, PG, SQDG, TAG, DGCC, DGDG, and DGTS_DGTA.

```{r Collinsplot}
#First, filter out everything that isn't what we want
species_to_keep <- c("PC", "PE", "PG", "SQDG", "TAG", "DGCC", "DGDG", "DGTS_DGTA")
collins_LOBdata <- filter(long_LOBdata, species%in%species_to_keep)

ggplot(data = collins_LOBdata) + 
  geom_point(aes(x=peakgroup_rt, y=peakgroup_mz, color=species)) +
  ylab("m/z") +
  xlab("Corrected retention time (min)") +
  scale_color_discrete(name="Lipid species") +
  theme(axis.title.y = element_text(face="italic"))
```

Looks good! Collins et al. also used the oxidation degree to add shapes, but
that's a good exercise for you to try on your own. (Hint: you'll have to add
an degree_oxidation column to long_LOBdata, and the ggplot argument is "shape")

We can also compare specific lipid species across the three treatments and 
watch how they change over time:

```{r Collins}
IPDAGs <- filter(long_LOBdata, lipid_class=="IP_DAG")

#Define custom factor levels so the facets and boxplots are in the right order
IPDAGs$treatment <- factor(IPDAGs$treatment, levels=c("0 uM", "30 uM", "150 uM"))
IPDAGs$time_point <- factor(IPDAGs$time_point, levels = c("4 hours", "8 hours", "24 hours"))

ggplot(data = IPDAGs) + geom_boxplot(aes(x=time_point, y=intensity, color=species)) +
  facet_wrap(~treatment)

#Better way to slice the data:
ggplot(data = IPDAGs) + geom_boxplot(aes(x=time_point, y=intensity, color=treatment)) +
  facet_wrap(~species, scales = "free_y") +
  theme(axis.text.x  = element_text(angle=90, vjust=1)) +
  ylab("Intensity") +
  xlab("Time sampled") +
  ggtitle("All IP-DAGs") +
  scale_color_discrete(name="Peroxide treatment")


ggplot(data = IPDAGs) + geom_boxplot(aes(x=treatment, y=intensity, color=time_point)) +
  facet_wrap(~species, scales = "free_y") +
  theme(axis.text.x  = element_text(angle=90, vjust=1)) +
  ylab("Intensity") +
  xlab("Peroxide treatment") +
  ggtitle("All IP-DAGs") +
  scale_color_discrete(name="Time sampled")

#What's that? An interesting subgroup analysis?

BLLs <- filter(long_LOBdata, species=="BLL")
ggplot(data = IPDAGs) + geom_boxplot(aes(x=treatment, y=intensity, color=time_point)) +
  theme(axis.text.x  = element_text(angle=90, vjust=1)) +
  ylab("Intensity") +
  xlab("Peroxide treatment") +
  ggtitle("BLL subgroup analysis") +
  scale_color_discrete(name="Time sampled")
```

Finally, we can use some of the code I wrote over the summer to get a better sense
for how the lipidome changes across these samples.

Heatmap of the lipidome across samples, grouped by lipid species:

```{r SURF}
fold_samps <- long_LOBdata %>%
  group_by(sample) %>%
  mutate(fold_change = log2(intensity/mean(intensity)))

ggplot(data = fold_samps, aes(x=species, y=sample)) + 
  geom_tile(aes(fill=fold_change), color="white") +
  scale_fill_gradient2(low = "darkorchid4", mid = "grey90", high = "darkgreen", na.value = "grey90") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ylab("Sample number") + xlab("Lipid Species")
```

Note that these haven't been normalized to DNPPE (the internal standard) - might
be another good thing to try doing on your own!

```{r SURFplot}
norm.rel.IPDAGs <- IPDAGs %>%
  group_by(sample) %>%
  mutate(proportion=(intensity/sum(intensity))*100)

#Aaaaand replot.
norm.rel.IPDAGs %>% group_by(sample) %>%
  ggplot(aes(x=sample, y=proportion)) + 
  geom_bar(aes(fill=species), stat = "identity") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ylab("Relative proportion") + xlab("Sample") +
  scale_fill_discrete(name="Lipid Species") +
  ggtitle("IP-DAGs in the PtH2O2lipids Dataset")
```

Et voila!
